---
title: "R Workshop (Code for Pittsburgh)"
author: "Conor Tompkins"
date: "4/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

##What is R
R is an interpreted programming language for statistics

##Intro to RStudio
RStudio is an Itegrated Development Environment for R
4 panels
-Scripts
-Console
-Environment
-Files

##How Does R Work?
```{r}
1
```

```{r}
1 + 2
```

```{r}
10 / 2
```

```{r}
5 * 2
```

```{r}
"this is a string. strings in R are surrounded by quotation marks."
```

```{r error = TRUE}
"1" + 1
```

```{r}
str(1)
str("1")
```




##Compare to Excel
R separates the data from the analysis. The data is stored in files (CSVs, JSON, etc). The analysis is stored in scripts. This makes it easier to share analysis performed in R. No need to take screenshots of your workflow in Excel. You have a record of everything that was done to the data.

##Compare to other programming languages

##Basic Functions
Functions to cover
-add
-subtract
-strings
-assign

##What is the Tidyverse?

##How does the Tidyverse work?
Tidyverse functions
-select
-filter
-mutate
-group_by
-summarize
-ggplot2

##Making graphs with 311 data
-load the data from WPRDC
-examine the data in R
-filter out non-pothole reports
-filter by date
-summarize total requests per month/year

##Make this graph in R
https://twitter.com/conor_tompkins/status/967028854652129281


```{r load_environment}
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(ggmap)
```

```{r load_data}
df <- read_csv("https://raw.githubusercontent.com/conorotompkins/pittsburgh_311/master/data/pittsburgh_311_2018_04_10.csv")

colnames(df) <- tolower(colnames(df))

df %>%
  mutate(date = ymd(str_sub(created_on, 1, 10)),
         time = hms(str_sub(created_on, 11, 18)),
         month = month(date, label = TRUE), 
         year = year(date),
         yday = yday(date)) -> df
```

```{r top_requests, message=FALSE, warning=FALSE, paged.print=FALSE, cache=TRUE}
df %>% 
  count(request_type, sort = TRUE) %>% 
  top_n(5) %>% 
  ungroup() -> df_top_requests
```

```{r area_chart, fig.height=6, fig.width=12, message=FALSE, warning=FALSE, cache=TRUE, paged.print=FALSE}
df %>% 
  semi_join(df_top_requests) %>% 
  count(request_type, month) %>% 
  ggplot(aes(month, n, group = request_type, fill = request_type)) +
  geom_area() +
  scale_fill_viridis("Request type", discrete = TRUE, option = "D") +
  scale_y_continuous(expand = c(0, 0),
                     labels = comma) +
  scale_x_discrete(expand = c(0, 0)) +
  labs(title = "Top 5 types of 311 requests in Pittsburgh",
       x = "",
       y = "Number of requests") +
  theme_bw(base_family = 18) +
  theme(panel.grid = element_blank())

```


##Make this graph in R

```{r map, fig.height=12, fig.width=12, message=FALSE, warning=FALSE, paged.print=FALSE, cache=TRUE}
df %>% 
  select(request_type, x, y) %>% 
  filter(!is.na(x), !is.na(y),
         request_type %in% c("Potholes", "Snow/Ice Removal")) -> df_map

city_map <-  get_map("North Oakland, Pittsburgh, PA", 
                     zoom = 12,
                     maptype = "toner", 
                     source = "stamen")

city_map <- ggmap(city_map)
```

```{r error = TRUE}
#Put the data on the map
city_map +
  stat_density_2d(data = df_map, #Using a 2d contour
                  aes(x, #longitude
                      y, #latitude
                      fill = request_type, #Use the count of arrests as the fill
                      alpha = ..level..), #Use alpha so you can see the map under the data
                  geom = "polygon") + #We want the contour in a polygon
  facet_wrap(~request_type) +
  scale_alpha_continuous(range = c(.1, 1)) +
  guides(fill = FALSE,
         alpha = guide_legend("Number of requests")) +
  labs(title = "The Two Seasons In Pittsburgh",
       subtitle = "311 requests in Pittsburgh",
       x = "",
       y = "") +
  theme_bw(base_family = 18) +
  theme(axis.text = element_blank())
```